---
title: "Unbalanced Rejection of Schengen Short Stay Visas"
author: "[Otho Mantegazza](https://othomantegazza.com/)"
date: today
editor_options: 
  chunk_output_type: console
---

# Background

This is an extract of [an analysis](https://blogs.lse.ac.uk/africaatlse/2023/06/22/europe-must-make-it-easier-for-africans-to-get-visas/) done by [Marta Foresti](https://uk.linkedin.com/in/marta-foresti-36255334) and me as part of the activity of the [LAGO Collective](https://www.lagocollective.org/) on how unfair and strict Schengen short stay visa policies might be hurting the development of the EU states, instead of protecting them.

This is a very short exploration of the same data done for teaching purposes.

# Introduction

The Schengen area grants [short stay visas](https://home-affairs.ec.europa.eu/policies/schengen-borders-and-visa/visa-policy_en) for visitors staying no longer than 90 days. And releases detailed statistics on the acceptance rate by consulate in which the request was lodged. Can we detect hint of an unbalance in those data?

# Analysis

## Packages and Setup

```{r}
#| warning: false
library(tidyverse)
library(readxl)
library(janitor)
library(countrycode)
library(paletteer)

theme_base <- 
    theme(
    plot.margin = margin(10,5,5,5),
    axis.ticks = element_line(),
    axis.line.y = element_line(),
    legend.position = "top",
    legend.box = "vertical"
  )

a <- theme_set(
  theme_minimal(
    base_size = 12,
  ) +
    theme_base
)
```

## Data

2022 statistics are available at the website for [Home Affairs of the European Commission](https://home-affairs.ec.europa.eu/policies/schengen-borders-and-visa/visa-policy_en) in excel format.

I've downloaded them manually and added them into the data folder.

Let's read and clean them:

```{r}
data_path <- 'data/Visa statistics for consulates in 2022_en.xlsx'

visa <-
  read_excel(
    data_path, 
    sheet = 2
  ) %>% 
  clean_names() %>%  
  select(
    schengen_state,
    consulate_country = country_where_consulate_is_located,
    consulate_city = consulate,
    issued = total_at_vs_and_uniform_visas_issued_including_multiple_at_vs_me_vs_and_lt_vs,
    not_issued = total_at_vs_and_uniform_visas_not_issued
  )
```

The visa dataset now looks like this:

```{r}
visa %>% glimpse()
```

## Missing Data

```{r}
visa %>% 
  summarise(
    across(
      .cols = everything(),
      .fns = ~is.na(.) %>% sum()
    )
  ) %>% 
  glimpse()
```

We can drop the observation with missing values in `consulate_country`, since they are not useful for this analysis.

```{r}
visa <- 
  visa %>% 
  drop_na(
    consulate_country
  )

visa %>% 
  summarise(
    across(
      .cols = everything(),
      .fns = ~is.na(.) %>% sum()
    )
  ) %>% 
  glimpse()
```


We can safely assume that the 'NA' in the column `issed` and `not_issued` are zeros instead. 

```{r}
visa <- 
  visa %>% 
  mutate(
    issued = issued %>% replace_na(0),
    not_issued = not_issued %>% replace_na(0)
  )

visa %>% 
  summarise(
    across(
      .cols = everything(),
      .fns = ~is.na(.) %>% sum()
    )
  ) %>% 
  glimpse()
```

## Recompute Statistics from Data

With the goal of visualization, we can recompute columns with totals and percentages of rejection.

```{r}
visa <- 
  visa %>% 
  mutate(tot_application = issued + not_issued,
         rej_rate = not_issued/tot_application)

visa %>% glimpse()
```


## Aggregate by Consulate Country

The consulate level is too detailed, let's try to visualize and then aggregate the data by country where the requesto was lodged:

```{r}
country_rank <- 
  visa %>% 
  group_by(consulate_country) %>% 
  summarise(
    mean_rej_rate = weighted.mean(
      x = rej_rate,
      w = tot_application)
  ) %>% 
  arrange(mean_rej_rate) %>% 
  pull(consulate_country)
```


```{r}
#| fig-height: 20
visa %>% 
  filter(tot_application > 0) %>% 
  ggplot() +
  aes(x = rej_rate,
      y = consulate_country %>% factor(levels = country_rank)) +
  geom_point(
    aes(size = tot_application),
    shape = 21,
    stroke = 1,
    fill = '#FFFF0077'
  ) +
  labs(title = 'Short Stay Visa Rejection Rate',
       x = 'Rejection rate [%]',
       y = 'Country where the application was lodged') +
  scale_radius(
    range = c(0, 5),
    limits = c(1, NA)
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, .05)),
    labels = scales::percent
  )
```

## Aggregate by Continent

Let's try to aggregate the countries by continent, to seek for patterns. We can infer the continent from the string `consulate_country` with functions from the package [countrycode](https://vincentarelbundock.github.io/countrycode/).

```{r}
visa <- 
  visa %>% 
  mutate(
    continent = consulate_country %>% 
      countrycode(
        origin = 'country.name',
        destination = 'continent'
      )
  )

visa %>% 
  glimpse()
```

And let's map the continent to the colour of the points.

```{r}
#| fig-height: 20
visa %>% 
  filter(tot_application > 0) %>% 
  ggplot() +
  aes(x = rej_rate,
      y = consulate_country %>% factor(levels = country_rank),
      colour = continent) +
  geom_point(
    aes(size = tot_application),
    alpha = .9
  ) +
  labs(title = 'Short Stay Visa Rejection Rate',
       x = 'Rejection rate [%]',
       y = 'Country where the application was lodged') +
  scale_radius(
    range = c(0, 5),
    limits = c(1, NA)
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, .05)),
    label = scales::percent
  ) +
  scale_color_paletteer_d(
    "awtools::mpalette"
  )
```

```{r}
#| fig-height: 10
visa %>%
  filter(tot_application > 0) %>% 
  ggplot() +
  aes(x = rej_rate,
      weight = tot_application,
      fill = continent) +
  geom_histogram() +
  geom_hline(yintercept = 0) +
  facet_wrap(facets = 'continent',
             ncol = 1) +
  labs(title = 'Short Stay Visa Rejection Rate',
       x = 'Rejection rate [%]',
       y = 'Application Lodged [n]') +
  scale_fill_paletteer_d(
    "awtools::mpalette"
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, .05)),
    label = scales::percent
  )
```


# Conclusions

Countries in the African continent face an unexpectedly high rejection rates for Schengen short stay visa applications.

This explorative analysis does not in any way explore the causes of this patterns, but highlights a potential problem that from here on can be described and studied more deeply.
